---
output:
  pdf_document: default
  html_document: default
---

# Probability Concepts

## Chapter Overview

In previous chapters, we discussed ways to describe variables and the relationships between them. From here, we want to start asking inferential statistics questions like "If my sample mean is 10, how likely is it that the population mean is actually 11?". Probability is going to start us on this path. 

Probability theory is the science of uncertainty and it is really interesting! But it can also be quite challenging. I try to frame probability around things most of us can do at home: flipping a coin, rolling a die, drawing from a deck of cards. You certainly don't need any of these things to get through this chapter, but you may find it helpful to have a coin/die/deck of cards on hand as you read through the examples. 

Take your time running practice problems and going through the examples, using a tactile approach like sorting through your deck of cards whenever it seems helpful. 

**Chapter Learning Objectives/Outcomes**

1. Find and interpret probabilities for equally likely events.
2. Find and interpret probabilities for events that are not equally likely.
3. Find and interpret joint and marginal probabilities.
4. Find and interpret conditional probabilities.
5. Use the multiplication rule and independence to calculate probabilities.

R Objectives: *none*

This chapter's outcomes correspond to course outcome (3) understand the basic rules of probability.

## Experiments, Sample Spaces, and Events

**Probability** is the science of uncertainty. When we run an experiment, we are unsure of what the outcome will be. Because of this uncertainty, we say an experiment is a **random process**. 

The probability of an event is the proportion of times it would occur if the experiment were run infinitely many times. For a collection of *equally likely events*, this looks like:
\[
  \text{probability of event} = \frac{\text{number of ways event can occur}}{\text{number of possible (unique) outcomes}}
\]

An **event** is some specified possible outcome (or collection of outcomes) we are interested in observing. 

> *Example*: If you want to roll a 6 on a six-sided die, there are six possible outcomes $\{1,2,3,4,5,6\}$. In general, we assume that each die face is equally likely to appear on a single roll of the die, that is, that the die is *fair*. So the probability of rolling a 6 is $$\frac{\text{number of ways to roll a 6}}{\text{number of possible rolls}} = \frac{1}{6}$$

> *Example*: We can extend this to a collection of events, say the probability of rolling a 5 or a 6: $$\frac{\text{number of ways to roll a 5 or 6}}{\text{number of possible rolls}} = \frac{2}{6}$$

The collection of all possible outcomes is called a **sample space**, denoted $S$. For the six-sided die, $S=\{1,2,3,4,5,6\}$.

To simplify our writing, we use **probability notation**:

- Events are assigned capital letters.
- $P(A)$ denotes the probability of event $A$.
- Sometimes we will also shorten simple events to just a number. For example, $P(1)$ might represent "the probability of rolling a 1".

We can estimate probabilities from a sample using a frequency distribution.

> *Example*: Consider the following frequncy distribution from section 1.6
>
> | Class | Frequency | 
> | :-: | :-: |
> | freshman | 12 |
> | sophomore | 10 |
> | junior | 3 |
> | senior | 5 |
>
> If a student is selected *at random* (meaning each student is equally likely to be selected), the probability of selecting a sophomore is $$\text{probability of sophomore} = \frac{\text{number of ways to select a sophomore}}{\text{total number of students}} = \frac{10}{30} \approx 0.3333$$ The probability of selecting a *junior or a senior* is $$\frac{\text{number of ways to select a junior or senior}}{\text{total number of students}} = \frac{3+5}{30} = \frac{8}{30} \approx 0.2667$$
>
> Using probability notation, we might let $A$ be the event we selected a junior and $B$ be the event we selected a senior. Then $$P(A \text{ or } B) = 0.2667$$

## Probability Distributions

Two outcomes are **disjoint** or **mutually exclusive** if they cannot both happen (at the same time). Think back to how we developed bins for histograms - the bins need to be nonoverlapping - this is the same idea!

> *Example*: If I roll a six-sided die one time, rolling a 5 and rolling a 6 are disjoint. I can get a 5 *or* a 6, but not both on the same roll.

> *Example*: If I select a student, they can be a freshman *or* a sophomore, but that student cannot be both a freshman and a sophomore at the same time. 

The outcome must be one event or the other (it cannot be both at the same time).

### Venn Diagrams

**Venn Diagrams** show events as circles. The circles overlap where events share common outcomes. 

```{r include=FALSE}
library(VennDiagram)
library(RColorBrewer)
```

```{r comment=NA}
venn.plot <- draw.pairwise.venn(
	area1 = 13,
	area2 = 12,
	cross.area = 3,
	category = c("Diamonds", "Face Cards"),
	fill = c("lightblue", "lightyellow"),
	lty = 1,
	cex = 2,
	cat.cex = 2,
	cat.pos = c(315, 20),
	cat.dist = c(0.09,0.09),
	cat.just = list(c(-1, -1), c(1, 1)),
	ext.pos = 30,
	ext.dist = -0.05,
	ext.length = 0.85,
	ext.line.lwd = 2,
	ext.line.lty = "dashed"
	)
grid.draw(venn.plot)
```

When a Venn Diagram has *no overlap* the events are mutually exclusive. This Venn Diagram shows the event "Draw a Diamond" and the event "Draw a Face Card". There are 13 diamonds and 12 face cards in a deck. In this case, the events are *not* mutually exclusive: it's possible to draw both a diamond and a face card at the same time: the Jack of Diamonds, Queen of Diamonds, and King of Diamonds. 

For quick reference, an image of a full 52-card deck is linked below. The "face cards" are the J, Q, and K. Each row represents a "suit". From top to bottom, the suits are clubs, spades, hearts, and diamonds. Cards can be either red (hearts and diamonds) or black (spades and clubs).

[Click here for a graphic of a standard 52 card deck.](https://www.cis.upenn.edu/~cis110/17fa/hw/hw08/standard52.jpg)

*On your own*: Consider events 

- $A$: "Draw a spade"
- $B$: "Draw a queen"
- $C$: "Draw a red"

Which of these events are mutually exclusive?

### Probability Axioms

A **probability distribution** lists all possible disjoint outcomes (think: all possible values of a variable) and their associated probabilities. This can be in the form of a table

| Roll of a six-sided die | 1 | 2 | 3 | 4 | 5 | 6 |
| :-: | :-: | :-: | :-: | :-: | :-: | :-: |
| Probability             | 1/6 | 1/6 | 1/6 | 1/6 | 1/6 | 1/6 |

(note that we could visualize this with a bar plot!) or an equation, which we will discuss in a later chapter. 

The **probability axioms** are requirements for a valid probability distribution. They are: 

1. All listed outcomes must be disjoint.
2. Each probability must be between 0 and 1. 
3. The probabilities must sum to 1. 

Note that #2 is true for ALL probabilities. If you ever calculate a probability and get a negative number or a number greater than 1, you know something went wrong!

>*Example*: Use the probability axioms to check whether the following tables are probability distributions.
>
> A)
>
> | X | {1 or 2} | {3 or 4} | {5 or 6} | 
> | :-: | :-: | :-: | :-: | 
> | P(X) | 1/3 | 1/3 | 1/3 |
>
> Each axiom is satisfied, so this is a valid probability distribution. 
>
> B) 
>
> | Y | {1 or 2} | {2 or 3} | {3 or 4} | {5 or 6} |
> | :-: | :-: | :-: | :-: | :-: |
> | P(Y) | 1/3 | 1/3 | 1/3 | -1/3 |
>
> In this case, the outcomes are not disjoint and one of the probabilities is negative, so this is *not* a valid probability distribution.

Probability distributions look a lot like relative frequency distributions. This isn't a coincidence! In fact, a relative frequency distribution is a good way to use data to approximate a probability distribution. 

### Exercises {-}
1. Use the probability axioms to determine whether each of the following is a valid probability distribution:
    A. 
    
    | x |	0	|	1	|	2	|	3	|	4 |	
    | :-: | :-: | :-: | :-: | :-: |
    | P(x)| 0.1|	0.2| 	0.1| 	0.3| 	0.1 |
    
    B. 
    
    | x	| 0 or 1| 1 or 2	| 3 or 4	| 5 or 6 |
    | :-: | :-: | :-: | :-: | :-: |
    | P(x)| 	0.1| 	0.2| 	0.4	| 0.3| 
2. Determine whether the following events are mutually exclusive (disjoint).
    a. Your friend studies in the library. You study at home.
    b. You and your study group all earn As on an exam. 
    c. You stay out until 3 am. You go to bed at 9 pm.
3. In a group of 24 people, 11 have cats and 13 have dogs. Four of them have both cats and dogs. Sketch a Venn Diagram for these events.

## Rules of Probability

Consider a six-sided die. $$P(\text{roll a 1 or 2}) = \frac{\text{2 ways}}{\text{6 outcomes}} = \frac{1}{3}.$$ Notice that we get the same result by taking $$P(\text{roll a 1})+P(\text{roll a 2}) = \frac{1}{6}+\frac{1}{6} = \frac{1}{3}.$$ It turns out this is widely applicable!

### Addition Rules 

***
<center><font size='4'><b>Addition Rule for Disjoint Outcomes</b></font></center>

If $A_1$ and $A_2$ are disjoint outcomes, then the probability that one of them occurs is $$P(A_1 \text{ or } A_2) = P(A_1)+P(A_2).$$ This can also be extended to more than two disjoint outcomes: $$P(A_1 \text{ or } A_2 \text{ or } \dots \text{ or } A_k) = P(A_1)+P(A_2)+\dots + P(A_k)$$ for $k$ disjoint outcomes.

***

Now consider a deck of cards. Let $A$ be the event that a card drawn is a diamond and let $B$ be the event it is a face card. (Check back to 3.2 for the Venn Diagram of these events.) 

- $A$: $\quad 2\diamondsuit$ $3\diamondsuit$ $4\diamondsuit$ $5\diamondsuit$ $6\diamondsuit$ $7\diamondsuit$ $8\diamondsuit$ $9\diamondsuit$ $10\diamondsuit$ $J\diamondsuit$ $Q\diamondsuit$ $K\diamondsuit$ $A\diamondsuit$
- $B$: $\quad J\heartsuit$ $Q\heartsuit$ $K\heartsuit$ $J\clubsuit$ $Q\clubsuit$ $K\clubsuit$ $J\diamondsuit$ $Q\diamondsuit$ $K\diamondsuit$ $J\spadesuit$ $Q\spadesuit$ $K\spadesuit$

The collection of cards that are diamonds or face cards (or both) is

<center>$\quad A\diamondsuit$ $2\diamondsuit$ $3\diamondsuit$ $4\diamondsuit$ $5\diamondsuit$ $6\diamondsuit$ $7\diamondsuit$ $8\diamondsuit$ $9\diamondsuit$ $10\diamondsuit$ $J\diamondsuit$ $Q\diamondsuit$ $K\diamondsuit$ $J\clubsuit$ $Q\clubsuit$ $K\clubsuit$ $J\heartsuit$ $Q\heartsuit$ $K\heartsuit$ $J\spadesuit$ $Q\spadesuit$ $K\spadesuit$</center>

Looking at these cards, I can see that there are 22 of them, so $$P(A \text{ or } B) = \frac{22}{52}$$

However, if I try to apply the addition rule for disjoint outcomes, $P(A)=\frac{13}{52}$ and $P(B)=\frac{12}{52}$ and I would get $\frac{13+15}{52} = \frac{25}{52}$, which isn't what we want! 

What happened? When I tried to add these, I *double counted* the Jack of Diamonds, Queen of Diamonds, and King of Diamonds (the cards that are in both $A$ and $B$). To deal with that, I need to subtract off the double count $\frac{13}{52}+\frac{12}{52}-\frac{3}{52}$. 

***
<center><font size='4'><b>General Addition Rule</b></font></center>

For any two events $A$ and $B$, the probability that *at least* one will occur is $$P(A \text{ or } B) = P(A)+P(B)-P(A \text{ and }B).$$

***

Notice that when we say "or", we include the situations where A is true, B is true, and the situation where are both A and B are true. This is an *inclusive or*. Basically, if I said "Do you like cats or dogs?" and you said "Yes." because you like cats *and* dogs, that would be a perfectly valid response. I recommend using the inclusive or with your friends any time you want to get out of making a decision. 

Also notice that the general addition rule applies to *any* two events, even disjoint events. This is because, for disjoint events, $P(A \text{ and } B) = 0$; it's impossible for both to occur at the same time!

### Complements

The **complement** of an event is all of the outcomes in the sample space that are *not* in the event. For an event $A$, we denote its complement by $A^c$.

> *Example*: For a single roll of a six-sided die, the sample space is all possible rolls: 1, 2, 3, 4, 5, or 6. If the event $A$ is rolling a 1 or a 2, then the complement of this event, denoted $A^c$, is rolling a 3, 4, 5, or 6. 
>
> We could also write this in probability notation: $S = \{1, 2, 3, 4, 5, 6\}$ and if $A=\{1,2\}$, then $A^c=\{3, 4, 5, 6\}$.

**Property**: $$P(A \text{ or } A^c)=1$$ Using the addition rule, $$P(A \text{ or } A^c) = P(A)+P(A^c) = 1.$$ Make sure you can convince yourself that $A$ and $A^c$ are *always* disjoint.

***
<center><font size='4'><b>Complement Rule</b></font></center>

$$P(A) = 1-P(A^c).$$

***


> *Example*: Consider rolling 2 six-sided dice and taking their sum. The event of interest is a sum less than 12. Find
>
> 1. $A^c$
> 2. $P(A^c)$
> 3. $P(A)$
>
> If $A =$ (sum less than 12), then $A^c =$ (sum greater than or equal to 12). Take a moment to notice that there is only one way to get a sum greater than or equal to 12: rolling two 6s. 
>
> The chart below shows the rolls of Die 1 as columns and the rolls for Die 2 as rows. The numbers in the middle are the sums. Note that there are 36 possible ways to roll 2 dice.
> 
> |  | 1 | 2 | 3 | 4 | 5 | 6 |
> | :- | :-: | :-: | :-: | :-: | :-: | :-: |
> | **1** | 2 | 3 | 4 | 5 | 6 | 7 |
> | **2** | 3 | 4 | 5 | 6 | 7 | 8 |
> | **3** | 4 | 5 | 6 | 7 | 8 | 9 |
> | **4** | 5 | 6 | 7 | 8 | 9 | 10 |
> | **5** | 6 | 7 | 8 | 9 | 10 | 11 |
> | **6** | 7 | 8 | 9 | 10 | 11 | 12 |
>
> Even without the chart, by noting that there's only one way to get a sum greater than or equal to 12, we can quickly find $P(A^C)$: $$ P(A^c) = \frac{1}{36}$$ 
> But trying to count all of the ways to get $A$ would take a long time! Instead, we can use $$P(A) = 1 - P(A^c) = 1-\frac{1}{36} = \frac{35}{36}$$

## Conditional Probability

A **contingency table** is a way to summarize **bivariate data**, or data from two variables. 

*Smallpox in Boston (1726)*
<table>
    <tbody>
        <tr>
            <td style="text-align: center;" colspan="2" width="130">
                <p>&nbsp;</p>
            </td>
            <td style="text-align: center;" colspan="3" width="254">
                <p style="text-align: center;"><strong>Inoculated</strong></p>
            </td>
        </tr>
        <tr>
            <td style="text-align: center;" colspan="2" width="130">
                <p>&nbsp;</p>
            </td>
            <td style="text-align: center;" width="74">
                <p><strong>yes</strong></p>
            </td>
            <td style="text-align: center;" width="84">
                <p><strong>no</strong></p>
            </td>
            <td style="text-align: center;" width="96">
                <p><strong>total</strong></p>
            </td>
        </tr>
        <tr>
            <td style="text-align: center;" rowspan="3" width="72">
                <p><strong>Result</strong></p>
            </td>
            <td style="text-align: center;" width="58">
                <p><strong>lived</strong></p>
            </td>
            <td style="text-align: center;" width="74">
                <p>238</p>
            </td>
            <td style="text-align: center;" width="84">
                <p>
                    <span style="color: #000000; background-color: #fbeeb8;">5136</span>
                </p>
            </td>
            <td style="text-align: center;" width="96">
                <p>
                    <span style="background-color: #bfedd2;">5374</span>
                </p>
            </td>
        </tr>
        <tr>
            <td style="text-align: center;" width="58">
                <p><strong>died</strong></p>
            </td>
            <td style="text-align: center;" width="74">
                <p>6</p>
            </td>
            <td style="text-align: center;" width="84">
                <p>844</p>
            </td>
            <td style="text-align: center;" width="96">
                <p>850</p>
            </td>
        </tr>
        <tr>
            <td style="text-align: center;" width="58">
                <p><strong>total</strong></p>
            </td>
            <td style="text-align: center;" width="74">
                <p>
                    <span style="background-color: #eccafa;">244</span>
                </p>
            </td>
            <td style="text-align: center;" width="84">
                <p>5980</p>
            </td>
            <td style="text-align: center;" width="96">
                <p>
                    <span style="background-color: #c2e0f4;">6224</span>
                </p>
            </td>
        </tr>
    </tbody>
</table>

<ul>
    <li>
        <span style="background-color: #fbeeb8;">5136</span> is the count of people who lived AND were not inoculated.&nbsp;
    </li>
    <li>
        <span style="background-color: #c2e0f4;">6224</span> is the total number of observations.
    </li>
    <li>
        <span style="background-color: #eccafa;">244</span> is the total number of people who were inoculated.
    </li>
    <li>
        <span style="background-color: #bfedd2;">5374</span> is the total number of people who lived.
    </li>
</ul>

This is basically a two-variable frequency distribution. And, like a frequency distribution, we can convert to proportions (relative frequencies) by dividing each count (each number) by the total number of observations:

<table>
    <tbody>
        <tr>
            <td style="text-align: center;" colspan="2" width="130">
                <p>&nbsp;</p>
            </td>
            <td style="text-align: center;" colspan="3" width="254">
                <p><strong>Inoculated</strong></p>
            </td>
        </tr>
        <tr>
            <td style="text-align: center;" colspan="2" width="130">
                <p>&nbsp;</p>
            </td>
            <td style="text-align: center;" width="74">
                <p><strong>yes</strong></p>
            </td>
            <td style="text-align: center;" width="84">
                <p><strong>no</strong></p>
            </td>
            <td style="text-align: center;" width="96">
                <p><strong>total</strong></p>
            </td>
        </tr>
        <tr>
            <td style="text-align: center;" rowspan="3" width="72">
                <p><strong>Result</strong></p>
            </td>
            <td style="text-align: center;" width="58">
                <p><strong>lived</strong></p>
            </td>
            <td style="text-align: center;" width="74">
                <p>0.0382</p>
            </td>
            <td style="text-align: center;" width="84">
                <p>
                    <span style="background-color: #fbeeb8;">0.8252</span>
                </p>
            </td>
            <td style="text-align: center;" width="96">
                <p>
                    <span style="background-color: #bfedd2;">0.8634</span>
                </p>
            </td>
        </tr>
        <tr>
            <td style="text-align: center;" width="58">
                <p><strong>died</strong></p>
            </td>
            <td style="text-align: center;" width="74">
                <p>0.0010</p>
            </td>
            <td style="text-align: center;" width="84">
                <p>0.1356</p>
            </td>
            <td style="text-align: center;" width="96">
                <p>0.1366</p>
            </td>
        </tr>
        <tr>
            <td style="text-align: center;" width="58">
                <p><strong>total</strong></p>
            </td>
            <td style="text-align: center;" width="74">
                <p>
                    <span style="background-color: #eccafa;">0.0392</span>
                </p>
            </td>
            <td style="text-align: center;" width="84">
                <p>0.9608</p>
            </td>
            <td style="text-align: center;" width="96">
                <p>
                    <span style="background-color: #c2e0f4;">1.0000</span>
                </p>
            </td>
        </tr>
    </tbody>
</table>
<ul>
    <li>
        <span style="background-color: #fbeeb8;">0.8252</span> is the proportion of people who lived AND were not inoculated.&nbsp;
    </li>
    <li>
        <span style="background-color: #c2e0f4;">1.000</span> is the proportion of total number of observations. Think of this as 100% of the observations.
    </li>
    <li>
        <span style="background-color: #eccafa;">0.0392</span> is the proportion of people who were inoculated.
    </li>
    <li>
        <span style="background-color: #bfedd2;">0.8634</span> is the proportion of people who lived.
    </li>
</ul>

The row and column totals are **marginal probabilities**. The probability of two events together ($A$ and $B$) is a **joint probability**.

What can we learn about the result of smallpox if we already know something about inoculation status? For example, given that a person is inoculated, what is the probability of death? To figure this out, we restrict our attention to the 244 inoculated cases. Of these, 6 died. So the probability is 6/244.

This is called **conditional probability**, the probability of some event $A$ if we know that event $B$ occurred (or is true): $$P(A|B) = \frac{P(A\text{ and }B)}{P(B)}$$ where the symbol | is read as "given".

> For death given inoculation, $$P(\text{death}|\text{inoculation}) = \frac{P(\text{death and inoculation})}{P(\text{inoculation})} = \frac{0.0010}{0.0392} = 0.0255.$$
> Notice that we could also write this as $$P(\text{death}|\text{inoculation}) = \frac{P(\text{death and inoculation})}{P(\text{inoculation})} = \frac{6/6224}{244/6224} = \frac{6}{244},$$ which is what we found when using the table to restrict our attention to only the inoculated cases.

If knowing whether event $B$ occurs tells us nothing about event $A$, the events are **independent**. For example, if we know that the first flip of a (fair) coin came up heads, that doesn't tell us anything about what will happen next time we flip that coin.

We can test for independence by checking if $P(A|B)=P(A)$. 

### Multiplication Rules

***
<center><font size='4'><b>Multiplication Rule for Independent Processes</b></font></center>
If $A$ and $B$ are independent events, then $$P(A \text{ and }B) = P(A)P(B).$$ 

***

We can extend this to more than two events: $$P(A \text{ and }B \text{ and } C \text{ and } \dots) = P(A)P(B)P(C)\dots.$$

Note that if $P(A \text{ and }B) \ne P(A)P(B)$, then $A$ and $B$ are *not* independent.

> *Example*: Find the probability of rolling a $6$ on your first roll of a die and a $6$ on your second roll.
>
> Let $A=$ (rolling a $6$ on first roll) and $B=$ (rolling a $6$ on second roll). For each roll, the probabiltiy of getting a $6$ is $1/6$, so $P(A) = \frac{1}{6}$ and $P(B)  = \frac{1}{6}$.
>
> Then, because each roll is independent of any other rolls, $$P(A \text{ and }B) = P(A)P(B) = \frac{1}{6}\times\frac{1}{6} = \frac{1}{36}$$

***
<center><font size='4'><b>General Multiplication Rule</b></font></center>
If $A$ and $B$ are any two events, then $$P(A \text{ and }B) = P(A|B)P(B).$$ 

***

Notice that this is just the conditional probability formula, rewritten in terms of $P(A \text{ and }B)$!

> *Example*: Suppose we know that 38.4\% of US households have dogs and that among those with dogs, 23.1\% have cats. Find the probability that a US household has both dogs and cats.
>
> Let $C=$ (household has cats) and $D=$ (household has dogs). We know from the problem statement that $P(D) = 0.384$. 
> 
> The other piece tells us something about the probability of having cats *among those with dogs*. This means that we *know* that these people have dogs. That is, *given* a household has dogs, the probabiltiy of cats is 23.1\%. In probability notation, $P(C|D) = 0.231$. Then $$P(C \text{ and }D) = P(C|D)P(D) = 0.231\times 0.384 = 0.0887$$ or the probability that a US household has both cats and dogs is 0.0887. 
